---
title: 'Level 1 Data Cleaning: Clean the mtcars Dataset'
---

```{r}
ls()
rm(list = ls())
```


# Objective

The objective of this assignment is to practice cleaning and transforming a messy dataset using tidyverse functions. You will use skills like renaming and reordering columns, sorting rows, changing data types, mutating data, and using the stringr and forcats packages.

In this Level 1 Cleaning assignment, you will work with a simple dataset and focus on basic data cleaning tasks. Most tasks are outlined in the assignment script. You may want to review the [Data Cleaning Walkthrough](https://github.com/nrdowling/d2mr-assessment/tree/main/01_data-cleaning/00_cleaning-walkthrough) before you begin.

You may additionally or alternatively complete the [Level 2 Data Cleaning assignment](https://github.com/nrdowling/d2mr-assessment/tree/main/01_data-cleaning/01_cleaning-level-2). In Level 2, you will work with a more complex dataset and perform additional cleaning tasks with less direct instruction. The Level 2 assignment has more opportunities to demonstrating meeting course standards than this Level 1 assignment and is recommended for those who are already comfortable with the tasks in this assignment.

# Instructions

1. If you have not already done so, pull the latest changes from the `d2mr-assessment` repository to ensure you have the most up-to-date version of the assignment files. Confirm you are working in your fork of the repository.
2. Open `cleaning-level-1.qmd` in RStudio and follow the instructions in the Setup section below to load and inspect the (clean) `goal.mtcars` dataset. 
    - **Important:** The provided code makes a small modification to the original `mtcars` dataset to create a `goal.mtcars` dataset. You will use this goal dataset as a reference to clean the `messy-mtcars.csv` dataset, not the true original.
3. Follow the tasks described in the assignment script to clean the dataset and return it to its original state.
4. At several points in this document you will come across questions or non-coding exercises. Answer these questions in the text of this .qmd document, immediately below the question.
5. *Optional:* Continue to follow the instructions in the assignment script to clean the dataset above and beyond matching the original. 


### Tasks

**Reminder:** Your goal is to match `goal.mtcars` (created below), *not* the default `mtcars`.[^1]

[^1]: Why? The `mtcars` dataset includes models names as *row names* rather than as a variable/column. For the purposes of this cleaning exercise, it's more useful to treat model names as a variable.

1. **Renaming Columns:** Rename columns to match the original `mtcars` dataset column names.
2. **Reordering Columns:** Ensure the columns are in the correct order.
3. **Sorting Rows:** Sort the rows by `mpg` (miles per gallon) and then by `cyl` (number of cylinders).
4. **Changing Data Types:** Ensure `gear` and `carb` are factors and other columns are numeric.
5. **Mutating Data:** Create a new column `hp_per_cyl` calculated as `hp` (horsepower) divided by `cyl`.
6. **String Manipulation:** Use `stringr` to clean any unwanted spaces or characters in the `model` column.
7. **Handling Factors:** Use `forcats` to ensure the levels of `gear` are ordered as `3`, `4`, `5`.

# Setup

## Loading libraries and set seed

```{r}
#| label: setup
library(tidyverse)
set.seed(1234)
```


## Read in and inspect messy data

Read in and inspect the messy dataset `messy-mtcars.csv`.

```{r}

#| label: read-messy-data

### LEAVE THIS CHUNK AS-IS ###

# You *might* need to edit the filepath, but don't change anything else!

# Read in messy-mtcars.csv
messy.mtcars <- read_csv(
  ########################################
  "messy-mtcars.csv", ## <-- THIS IS THE ONLY THING IN THIS CHUNK YOU CAN CHANGE IF NECESSARY
  ########################################
  trim_ws = FALSE, name_repair = "minimal", col_types = cols(.default = col_character()))

# Inspect the data
head(messy.mtcars)

```

## Inspect the original mtcars dataset

```{r}
#| label: inspect-original-data

### LEAVE THIS CHUNK AS-IS ###

# Load the original mtcars dataset
data(mtcars)

# Create the "goal.mtcars" dataset
# Convert row names to a column called "model" (see note above about row names)
goal.mtcars <- mtcars %>%
  rownames_to_column(var = "model")

# Inspect the goal.mtcars dataset
head(goal.mtcars)

# Optionally inspect the original mtcars dataset to see what the row names vs column issue looks like
# head(mtcars)

```


QUESTIONS:

1. What are the differences between the messy dataset and the original mtcars dataset?

<!-- answer below --> 
The messy dataset has two extra columns compared to the original mtcars dataset: an extra dsip column and a "wt-run if" column. The numbers in the original database are rounded to 3 decimals at the most, whereas they are much longer in the messy dataset. There are also variations in the names of both the rows and columns. The messy data also has all of its columns set at character vectors containing strings (<chr>) versus the original set which has its columns set at double vectors containing real numbers (<dbl>). Also the columns are out of order.

2. What are the main issues you need to address in cleaning?

<!-- answer below --> 
The columns should all match in name, order, and data type.



# Clean the Dataset

## Creat "clean" dataset

```{r}

#| label: make-cleaning-dataset

# Create a dataset to work with during the cleaning process called "clean.mtcars"
clean.mtcars <- messy.mtcars
```


## Clean columns/variables


```{r}
library(dplyr)

colnames(messy.mtcars)

#| label: rename-columns

# Rename columns to match the original mtcars dataset
clean.mtcars <- messy.mtcars %>%
    # The format is rename(new_name = old_name)
  rename(model = ` Car Model`, mpg = `mpg (Miles/Gallon)`, cyl = cylinders, disp = dsip, hp = horsepower, gear = gearbox, carb = carburetors)
```

```{r}

colnames(clean.mtcars)
colnames(goal.mtcars)

#| label: reorder-columns

# Reorder columns to match the original mtcars dataset
clean.mtcars <- clean.mtcars[, c("model", "mpg", "cyl", "disp","hp", "drat", "wt", "qsec", "vs", "am", "gear", "carb")]

#| label: compare-column-names-2

# Compare the column names of clean.mtcars and goal.mtcars
all.equal(colnames(clean.mtcars), colnames(goal.mtcars))

```

```{r}
colnames(clean.mtcars)
#| label: correct-data-types

# Correct variable data types
clean.mtcars<- clean.mtcars %>% 
    # Convert to numeric
    mutate(
        mpg = as.numeric(mpg),
        cyl = as.numeric(cyl),
        disp = as.numeric(disp),
        hp = as.numeric(hp),
        drat = as.numeric(drat),
        wt = as.numeric(wt),
        qsec = as.numeric(qsec),
        vs = as.numeric(vs),
        am = as.numeric(am),
        gear = as.numeric(gear),
        carb = as.numeric(carb)
    ) %>%
#| label: round-numeric-values
# Round numeric values
    # Round to correct decimal places
    mutate(
        mpg = round(mpg, 1),
        disp = round(disp, 1),
        drat = round(drat, 2),
        wt = round(wt, 3),
        qsec = round (qsec, 2)
    )


```

## Checkpoint 1

```{r}

#| label: checkpoint-1

# Inspect the current state of the dataset

# Use all.equal() to check if clean.mtcars matches goal.mtcars
all.equal(clean.mtcars, goal.mtcars)
render_diff(diff_data(clean.mtcars, goal.mtcars))

# Inspect the goal.mtcars dataset and compare to clean,mtcars
head(clean.mtcars)
head(goal.mtcars)

# Check the class of each dataset
class(clean.mtcars)
class(goal.mtcars)
```

Questions:

3. The current state of the dataset does not yet match the `goal.mtcars` dataset. Explain the issues that `all.equal()` flags.  Are there groups of issues that can be addressed together?

<!-- answer below -->
The class attribute (e.g., "data.frame", "list", "tibble") differs between clean.mtcars and goal.mtcars. The clean.mtcars has three classes, while the goal.mtcars has only the one "data.frame". As such, there is a mismatch in the class attribute, likely because the two objects are of different types (e.g., a data frame vs. a tibble). These issues can be addressed together.

Additionally, there are 13 mismatches in the "model" column due to different values, missing values, or ordering differences between the two data sets. The "cyl" and "hp" columns also have mismatches in that the clean.mtcars is missing values. These issues can likely be addressed together.

3. Install (if necessary) and load the `daff` package. In the code chunk below, use the `diff_data()` function to identify the differences between `clean.mtcars` and `goal.mtcars`. 

```{r}
#| label: diff-data

# Install and/or load the daff package if needed
# install.package("daff")
# library(daff)
library(daff)

# Use render_diff() and diff_data() to identify differences between clean.mtcars and goal.mtcars
diff_data(clean.mtcars, goal.mtcars)
render_diff(diff_data(clean.mtcars, goal.mtcars))

```

How is this method of identifying data discrepancies different from `all.equal()`? Which do you find more helpful at this point?

<!-- answer below --> 
Whereas the 'all.equal()' method gives general differences between data sets, the diff-data method gives specifics. Given how many discrepancies there are at the moment, I find the general information from all.equal to be more helpful at this point.

## Clean data values

Identifying specific value problems and cleaning them might take a little creativity at times! Use the troubleshooting/debugging workflows we've discussed to get as far as you can.

```{r}

#| label: remove-unwanted-spaces

# Remove unwanted whitespace in any columns required
clean.mtcars <- clean.mtcars %>%
    mutate(
        # Trim whitespace from the beginning and end of the strings
        model = str_squish(model),
        mpg = str_squish(mpg),
        cyl = str_squish(cyl), 
        disp = str_squish(disp),  
        hp = str_squish(hp),
        drat = str_squish(drat),  
        wt = str_squish(wt),
        qsec = str_squish(qsec), 
        vs = str_squish(vs),  
        am = str_squish(am),
        gear = str_squish(gear),
        carb = str_squish(carb),
        )
```

```{r}

#| label: impute-missing-values

# Impute missing values for cyl and hp

# Manually impute missing values in column 'cyl'
clean.mtcars$cyl[is.na(clean.mtcars$cyl)][1] <- 6  # Impute the first NA with 6

# Manually impute missing values in column 'hp'
clean.mtcars$hp[is.na(clean.mtcars$hp)][1] <- 66 # Impute the first NA with 66


```


```{r}

#| label: fix-typos

# Fix typos in model names
## Hint: This can be completed with a 7-line mutate() statement
clean.mtcars <- clean.mtcars %>% 
    # Replace typos with corrections
    mutate(
        model = str_replace_all(model, "Hornet SportAbout", "Hornet Sportabout"),
        model = str_replace_all(model,"CAdillac Fleetwood", "Cadillac Fleetwood"),
        model = str_replace_all(model,"ToyotA Corolla", "Toyota Corolla"),
        model = str_replace_all(model,"TOyota Corona", "Toyota Corona"),
        model = str_replace_all(model,"PontiAc Firebird", "Pontiac Firebird"),
        model = str_replace_all(model,"L0tus Europa", "Lotus Europa"),
        model = str_replace_all(model,"FerrAri Dino", "Ferrari Dino")
    ) 
  

```

```{r}

#| label: remove-extra-column

# Remove extra column 'hp_per_cyl'
clean.mtcars <- clean.mtcars %>% select(-hp_per_cyl)


```


## Checkpoint 2

```{r}
#| label: checkpoint-2

# Inspect the dataset and compare to goal.mtcars
head(clean.mtcars)
head(goal.mtcars)

# all.equal()
all.equal(clean.mtcars, goal.mtcars)
render_diff(diff_data(clean.mtcars, goal.mtcars))

# daff
#| label: diff-data

# Install and/or load the daff package if needed
# install.package("daff")
# library(daff)
library(daff)

# Use render_diff() and diff_data() to identify differences between clean.mtcars and goal.mtcars
diff_data(clean.mtcars, goal.mtcars)
render_diff(diff_data(clean.mtcars, goal.mtcars))

```

Questions:

5. Is your dataset identical to `goal.mtcars`? If not, what are the remaining issues? If there were any issues you could not resolve in code, describe a proposed solution in plain English.

<!-- answer below -->
The datasets are not identical. There is still a difference in class across the datasets, the model column still has a mismatch, and there is a mean difference between the values in hp across the two datasets. First, I will make both datesets into tibbles. Then, I will check the model column for mismatches and then make the necessary fixes. Finally, I will compare the summary statistics of the hp column in both datasets and see if there is a clear way to fix the discrepancy.


```{r}

#| label: make-tibbles

# Make both datasets tibbles
clean.mtcars <- as_tibble(clean.mtcars) 
# this already was a tibble, but I'll run it again to make sure
goal.mtcars <- as_tibble(goal.mtcars)


#| label: find-correct-models-mismatch

# find where the mismatch exists in model column
unique(clean.mtcars$model)

# fix the mismatch by replacing typo with correction
clean.mtcars <- clean.mtcars %>% 
    # Replace typos with corrections
    mutate(
        model = str_replace_all(model,"T0yota Corona", "Toyota Corona")
    )


#| label: hp-summary-statistics-mismatch

# look at the summary statistics for hp column in clean.mtcars
summary(clean.mtcars$hp)
# look at the summary statistics for hp column in goal.mtcars
summary(goal.mtcars$hp)

# Find rows where hp values differ between clean.mtcars and goal.mtcars
mismatches <- clean.mtcars$hp != goal.mtcars$hp
which(mismatches)  # Get the indices of mismatched rows

#| label: hp-column-correction

# correct value in row 27 of hp column so that datasets match
clean.mtcars$hp[clean.mtcars$model == "Porsche 914-2"] <- 91  # Replace with correct value

```


## Checkpoint 3


```{r}
#| label: checkpoint-3

# Inspect the dataset and compare to goal.mtcars
head(clean.mtcars)
head(goal.mtcars)

# all.equal()
all.equal(clean.mtcars, goal.mtcars)
render_diff(diff_data(clean.mtcars, goal.mtcars))

# daff
#| label: diff-data

# Install and/or load the daff package if needed
# install.package("daff")
# library(daff)
library(daff)

# Use render_diff() and diff_data() to identify differences between clean.mtcars and goal.mtcars
diff_data(clean.mtcars, goal.mtcars)
render_diff(diff_data(clean.mtcars, goal.mtcars))
```

# Optional Cleaning Tasks

You can call it a day here, or continue with the optional cleaning tasks below. These tasks are a chance to challenge yourself and practice your skills further. They are also additional opportunities to demonstrate mastery of course standards! 


## Additional guided cleaning

*Optional:* Complete any number of the following tasks to further clean the dataset. Each task should be in its own code chunk with an appropriate label and clearly commented code:

1. Create a new factor column `mpg_category` that categorizes `mpg` into "low", "medium", and "high" based on the distribution of `mpg` values.

```{r}

#| label: mpg-category
# Create mpg_category column
clean.mtcars <- clean.mtcars %>%
  mutate(mpg_category = cut(mpg,
                            breaks = quantile(mpg, probs = c(0, 0.33, 0.67, 1), na.rm = TRUE),
                            labels = c("low", "medium", "high"),
                            include.lowest = TRUE))

```

2. Create a new factor column `wt_category` that categorizes `wt` into "light", "medium", and "heavy" based on the distribution of `wt` values, then reverse the order of the levels.

```{r}

#| label: wt-category-reversed
# Categorize wt and reverse levels
clean.mtcars <- clean.mtcars %>%
  mutate(wt_category = cut(wt,
                           breaks = quantile(wt, probs = c(0, 0.33, 0.67, 1), na.rm = TRUE),
                           labels = c("light", "medium", "heavy"),
                           include.lowest = TRUE),
         wt_category = fct_rev(wt_category))  # Reverse the levels

```


3. Create a new column `mpg_per_cyl` that calculates `mpg` divided by `cyl`.

```{r}
#| label: mpg-per-cyl
# Create mpg_per_cyl column
clean.mtcars <- clean.mtcars %>%
  mutate(mpg_per_cyl = mpg / cyl)

```


4. Filter the data to only include cars with automatic transmissions. (Hint: `?mtcars` will show you useful info.)

```{r}
#| label: filter-am
# Filter for cars with automatic transmissions
automatic_cars <- clean.mtcars %>%
  filter(am == 0)

```


5. Identify which variable would be better suited as a logical variable, then convert it to logical.

```{r}
#| label: identify-logical-variable
# Identify columns with exactly two unique values
binary_cols <- clean.mtcars %>%
  summarise(across(everything(), ~ n_distinct(.) == 2)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "is_binary") %>%
  filter(is_binary)

binary_cols

#| label: convert-logical-variables
# Convert am to logical variable
clean.mtcars <- clean.mtcars %>%
  mutate(am_logical = as.logical(am))  # Convert `am` to logical

# Convert vs to logical variable
clean.mtcars <- clean.mtcars %>%
  mutate(vs_logical = as.logical(vs))  # Convert `am` to logical
```


6. Sort the dataset by `mpg_category` then reverse alphabetically by `model` (so that models with the same `mpg_category` are sorted Z to A).

```{r}
#| label: sort-dataset
# Sort by mpg_category (factor order) and reverse alphabetically by model
clean.mtcars <- clean.mtcars %>%
  arrange(mpg_category, desc(model))

```


7. Write the cleaned dataset as a csv file called `clean-mtcars.csv`, then read the csv back in maintaining the correct data types.


## Unguided cleaning and transformation

*Optional:* If you have the time and interest, continue transforming this dataset as you please. Create new columns based on the existing ones, reformat strings, try your hand at a regex replacement, summarize by groups (factor levels), visualize a simple relationship, or anything else you can think of. You can do this in addition to or instead of the suggested additional cleaning tasks above.


# Submission & Assessment

To submit:

1. Add & modify the `assessment.md` in this mini-project's directory:
    1. Check off all objectives you believe you have demonstrated
    2. Indicate which objectives you are meeting for the first time (if any)
    3. Complete any relevant open-ended items
2. Push your changes to your centralized assignment repository on GitHub. 
3. Confirm that Dr. Dowling and your section TA are added as collaborators to your repository.
4. Submit your work in your next open mini-project assignment by including the following information in the text box:
    1. The title of the assignment: "Level 1 Data Cleaning: Clean the mtcars Dataset"
    2. A link to the **directory** for this assignment in your centralized assignment repo


